
\documentclass[a4paper, 11pt, fleqn]{book}

% Set equal margins on book style
\setlength{\oddsidemargin}{53pt}
\setlength{\evensidemargin}{53pt}
\setlength{\marginparwidth}{57pt}
\setlength{\footskip}{30pt}

\begin{document}

\title{Objective-C Run-time}
\author{Bc. Kry\u{s}tof V\'{a}\u{s}a}
\date{}
\maketitle

\chapter{Objective-C}
\section{Foreword}

There are slight differences between the iOS and OS X versions of the run-time (e.g. iOS doesn't support garbage collection and only the new 2.0 run-time is available). Within this thesis, when talking about Apple's implementation of the run-time, the OS X version will be the one talked about.

Even though I will mention a few words about the garbage collection and ARC\footnote{ARC - automatic reference counting, a feature introduced in Xcode 4.2 (Xcode is Apple's IDE) that uses compiler's static analysis combined with compiler annotations.}, not much attention will be paid to it as garbage collection is being deprecated in OS X 10.8 (and has severe dependencies on Mac OS X itself) and ARC is relatively new and uses a lot of compiler-dependent features as well as auto-zeroing weak links, etc.; which is beyond the scope of this work. Instead, the focus will be put on general functionality of the run-time, analysis of the current implementations and designing the modular run-time itself.

\section{What is Objective-C?}

  Objective-C is an object-oriented programming language that is a strict superset of C. Any C code can be used within Objective-C source code. Its run-time is written in C as well, some parts in assembly language or more recently in C++ (more about that later on).
  
\section{Brief history of Objective-C}

In the early 1980s, Brad Cox and Tom Love, at their company Stepstone, inspired by the dynamic nature of Smalltalk decided to bring object-oriented concept to the world of C while maintaining full backward compatibility.

In 1988, NeXT has licensed Objective-C from Stepstone, added Objective-C support to the GCC compiler and decided to use it in its OpenStep and NeXTStep operating systems.

After Apple had acquired NeXT in 1996, Objective-C stayed alive in Rhapsody and later on Mac OS X, where it's the main programming language to the date.

For this whole time, the Objective-C language stayed almost the same without any big changes. In 2006, Apple announced Objective-C 2.0 (which was release in Mac OS X 10.5 in 2007), which introduced garbage collection (which is deprecated in 10.8, however, in favor to more efficient ARC - automatic reference counting - see below \footnote{http://cocoaheads.tumblr.com/post/17719985728/10-8-objective-c-enhancements}), properties (object variables with automatically generated getters and/or setters with specified memory management), fast enumeration (enumeration over collections in a foreach-style), and some other minor improvements.

Lately, more improvements have been made to Objective-C, most importantly ARC (automatic reference counting). Apple's run-time assumes some method names to be special. In particular, -retain, -autorelease, -release methods are used for memory management, which (unless garbage collection is turned on) handle the reference count for the particular object. ARC requires compiler support, though - all it does is that it automatically inserts -retain/-release/-autorelease (together called ARR) calls into the code (can be modified using \verb=__attribute__= construct) and automatically generates a -dealloc method (which is called when the object is being deallocated).

This, however, represents a problem - none of the ARR calls must be called directly in the code - hence you need to convert all of your code to ARC. One disadvantage which results in a big advantage - compatibility with all libraries (Apple calls Objective-C libraries frameworks) - this was a big disadvantage of garbage collection: 

You could keep the code as it was as the run-time itself redirected the ARR methods to a no-op function on the fly, however, all linked libraries/frameworks/plugins needed to be recompiled with garbage collection support turned on. This caused two things: mess in the code as if you migrated your code to garbage-collection-enabled environment, it was riddled with ARR calls, however, newly written code typically omitted those calls, making the code inconsistent; and some libraries never got GC support anyway.

In the forthcoming (TODO: at the time this will be done, it will be already released) release of OS X 10.8, several new features will be included - default synthesis of getters (in prior versions, you had to declare \verb=@property= in the header file and use \verb=@sythesize= or \verb=@dynamic= in the implementation file - see Syntax of Objective-C), type-safe enums, literals for NSArray, NSDictionary and NSNumber (classes declared in Apple's Foundation framework), etc.

\section{Compilation of Objective-C}

As Objective-C is a strict superset of C, as mentioned above, all C code is a valid Objective-C code. This thesis assumes that you have some brief knowledge of this language, at least syntax-wise.

This section will cover compilation of Objective-C and run-time's involvement in this process.

\subsection{Creating classes in code}

It is possible to dynamically create classes while the application is running, add methods, properties, implemented protocols, etc. to them. Actually all of Objective-C code could be translated to plain C\footnote{There's a LLVM Clang compiler option -rewrite-objc which will convert all the Objective-C syntax into calls of pure C methods - the run-time methods. If I run clang -rewrite-objc test.m, where test.m contains the Objective-C code, a new test.cpp (yes, it's actually a C++ file, though no C++ features are used) is created, containing the translated code.}. Here's an example:


This is a sample code that calls two methods - each on a different object, though\footnote{As will be explained later on, each class actually consists of two classes - the meta class, which has the +alloc method and the regular class, which has the -init method.}:
\begin{verbatim}SomeClass *myObj = [[SomeClass alloc] init];\end{verbatim}

- will be translated to:

\begin{verbatim}SomeClass *myObj = ((id (*)(id, SEL, ...))(void *)objc_msgSend)
((id)((id (*)(id, SEL, ...))(void *)objc_msgSend)
(objc_getClass("SomeClass"), sel_registerName("alloc")), 
sel_registerName("init"));\end{verbatim}

- which after removing the casting and adding a little formatting is:

\begin{verbatim}SomeClass *myObj = 
  objc_msgSend(
    objc_msgSend(
      objc_getClass("SomeClass"),  
      sel_registerName("alloc")), 
    sel_registerName("init"));
\end{verbatim}

So it's two nested \verb=objc_msgSend= calls. \verb=objc_msgSend= is a method that can be said to be the core of Objective-C run-time. It's the most used function of the run-time. Every method call in Objective-C gets translated into this variadic function call, which takes \verb=self= as the first argument (i.e. the object that the method is called on, or the message is sent to\footnote{Here can be seen the Smalltalk influence.}), the second argument is a selector (generally the method's name) and can be followed by arguments.

The run-time then looks up the object's class, finds a function that implements this method (so called \verb=IMP=) and calls it. There's several things to point out:

• method \emph{names} are used. \verb=sel_registerName= is a function that makes sure that for that particular method name only one selector pointer is kept.
• each of the calls goes to a different object. The first call gets to something returned by \verb=objc_getClass= which is a returns an instance of a meta class
• every class consists of two classes - one regular of which you create objects and one meta - which typically (unless you manually craft another one) has only one instance: a receiver for class methods (static methods).

If you care to investigate this hierarchy, here's a small example:

\verb=NSObject= class is part of the Foundation framework Apple supplies. Even though its often assumed to be the one and only root class in Objective-C. This is quite incorrect: there can be as many root classes in Objective-C as you wish - \verb=NSProxy= is an example and you can easily create your own.

\begin{verbatim}@interface ClassWithoutSuperclass
@end\end{verbatim}

This will declare a new root class. It has absolutely no functionality - no memory management \verb=retain= and \verb=release=, no \verb=+alloc= method is declared either - you wouldn't be able to even create a new instance of this class without the run-time function \verb=class_createInstance=\footnote{This is basically why it is recommended to inherit everything from NSObject (or any other basic object class) which implements some basic communication with the run-time as well as some basic memory management, etc.}.

Let's use this simplified class structure for representing an object (it's obviously more complex in real life):

\begin{verbatim}
typedef struct class_t {  struct class_t *isa;  struct class_t *superclass;
} objc_class_t;
\end{verbatim}

Every object is a pointer to a structure like this one. The first field, so-called \verb=isa=, is a pointer to the class structure (to the meta class instance). Second field points to the superclass. It might be confusing at first that it's a class structure, but remember that even the object's class is actually an object - an instance of the meta class\footnote{The root meta class's isa pointer points to the structure itself - i.e. there's a pointer cycle.}.

In case of ClassWithoutSuperclass:

\begin{verbatim}
objc_class_t *cl_str = objc_getClass("ClassWithoutSuperclass");

=============
cl_str                    0x1000054d8
cl_str->isa               0x100005500
cl_str->superclass        0x0

cl_str->isa->isa          0x100005500
cl_str->isa->superclass   0x1000054d8
\end{verbatim}

As you can see, there becomes a cycle.

A few more syntax is introduced, however. Basic method called on an object is placed between [ and ] brackets:

\begin{verbatim}
  [cat eatMice];
\end{verbatim}

If you desire to pass an argument, it is passed after a semi-colon:

\begin{verbatim}
  [cat eat:mouse];
\end{verbatim}

If more arguments are desired, the method's name splits into parts:

\begin{verbatim}
  [cat eat:mouse andDrink:milk];
\end{verbatim}

To declare a class, use the \verb=@interface= construct:

\begin{verbatim}
@interface SomeClass {

}

+(void)doSomethingStatic;
-(void)secondMethod:(void*)firstArgument withTwoArguments:(int)secondArgument;
-(void)someMethod:(void*)firstArgument;
@end

\end{verbatim}

Between the { } brackets, you may place variables, like into a C struct. This class is a root class, i.e. doesn't inherit from any other class. In Objective-C code, your classes will usually inherit from \verb=NSObject= class if Apple's frameworks are used. This can be declared as \verb=@interface SomeClass : NSObject=.

If you need to declare protocols implemented (protocols are similar to Java's interfaces - a list of methods declared between \verb=@protocol ProtocolName= and \verb=@end=), add a list of protocols between \verb=<= and \verb=>= brackets:

\begin{verbatim}
  @interface SomeClass : NSObject <ProtocolName, OtherProtocol>
\end{verbatim}

In Objective-C 2.0, you can take advantage of properties. To declare a property, use the \verb=@property= construct:

\begin{verbatim}
  @property (readwrite, assign, getter=isHealthy) BOOL healthy;
\end{verbatim}

The first argument can be either \verb=readwrite= or \verb=readonly=, second can be either \verb=assign= (default for non-object values), \verb=retain= (default for objects) or \verb=copy= which copies the object. Note that this doesn't mean any automatic object copying - the object must then implement the \verb=NSCopying= protocol, which defines -copy and -copyWithZone: methods. The last \verb=getter= argument can be used for modifying the getter name - without the third argument, this declaration would yield in -healthy and -setHealthy: method declarations. With the third argument, -isHealthy is generated (note that the convention is not to prefix getters with 'get'). 

These methods may be completely omitted if the so-called dot-syntax is used. It may feel familiar to Java or C# developers, accessing fields using the '.' - \verb=BOOL h = obj.healthy;= and \verb=obj.healthy = h;=, however, my personal stance is against the dot-syntax - as Objective-C is a strict superset of C, the dot-syntax should (in my opinion) be used only for structs, not objects which are allocated on the heap (unless some special workarounds are used\footnote{http://www.mikeash.com/pyblog/friday-qa-2010-01-15-stack-and-heap-objects-in-objective-c.html}). The dot-syntax gets translated into regular Objective-C calls during compilation anyway.

Method declarations begin with either \verb=-= or \verb=+= - the \verb=-= is used for instance methods, \verb=+= is used for class (static) methods.


\begin{verbatim}

@implementation SomeClass
+(void)doSomethingStatic{
  // Nothing to do
}
-(void)secondMethod:(void*)firstArgument withTwoArguments:(int)secondArgument{
    // Do something meaningful, at last!
}
-(void)someMethod:(void*)firstArgument{
    [self secondMethod:firstArgument withTwoArguments:1];
}
@end

\end{verbatim}

The method implementations are written between \verb=@implementation SomeClass= and \verb=@end=. The rest is just like in regular C with functions - the declaration of the method is followed by the code itself which is enclosed in { and } brackets.

If you have declared \verb=@property='s in the class' interface (typically placed in the header file), you need to (unless you are using OS X 10.8 or later), use either \verb=@synthesize= or \verb=@dynamic=. \verb=@synthesize= is followed by a list of properties to synthesize. Optionally you can specify onto which object variable is the property mapped. For example,

\begin{verbatim}
  @synthesize healthy = _myVariable;
\end{verbatim}

will take \verb=someProperty= and map it to \verb=_myVariable=, i.e. calling \verb=[obj setHealthy:YES]= will result in \verb=_myVariable= being \verb=YES= (as C doesn't have native boolean type, the run-time defines \verb=BOOL= as \verb=unsigned char= and two values \verb=YES (1)= and \verb=NO (0)=).

Using \verb=@dynamic= tells the compiler not to generate any methods, but you implement them on your own (for that particular property).


\section{Run-time of Objective-C}

As has been mentioned above, Objective-C is a strict superset of C language, where all objects are pointers to allocated structs with a specific structure (e.g. the first field of the struct is so-called isa - a pointer to the class of that particular object).
  
All of the Objective-C code can be actually rewritten in plain C - there's a LLVM Clang compiler option -rewrite-objc which will convert all the Objective-C syntax into calls of pure C methods - the run-time methods. If I run \begin{verbatim}clang -rewrite-objc test.m\end{verbatim}, where test.m contains the code that has been presented in the Syntax of Objective-C section, a new test.cpp (yes, it's actually a C++ file, though no C++ features are used) is created, containing the following code \footnote{I have taken the liberty of formatting the code and shortening it a little - it contains external run-time method and struct declarations, so that you can actually compile the file with a C compiler - provided you're linking against the Objective-C run-time library. Also, at the end, there were some linker-specific declarations.}:

\begin{verbatim}
typedef struct objc_object SomeClass;

static void _C_SomeClass_doSomethingStatic(Class self, SEL _cmd) {
  // Nothing to do
}

static void _I_SomeClass_secondMethod_withTwoArguments_
  (SomeClass * self, SEL _cmd, void *firstArgument, 
  int secondArgument) {
  // Do something meaningful, at last!
}

static void _I_SomeClass_someMethod_(SomeClass * self, SEL _cmd, 
  void *firstArgument) {
  ((void (*)(id, SEL, void *, int))(void *)objc_msgSend)((id)self,
     sel_registerName("secondMethod:withTwoArguments:"), 
     (void *)firstArgument, 1);
}

static struct {
   struct _objc_method_list *next_method;
   int method_count;
   struct _objc_method method_list[2];
} _OBJC_INSTANCE_METHODS_SomeClass 
  __attribute__((used, section ("__OBJC, __inst_meth")))= {
  0, 
  2,
  {
    {
      (SEL)"secondMethod:withTwoArguments:",
      "v28@0:8^v16i24",
      (void *)_I_SomeClass_secondMethod_withTwoArguments_
    }
    {
      (SEL)"someMethod:",
      "v24@0:8^v16",
      (void *)_I_SomeClass_someMethod_
    }
  }
};

static struct {
	struct _objc_method_list *next_method;
	int method_count;
	struct _objc_method method_list[1];
} _OBJC_CLASS_METHODS_SomeClass 
  __attribute__((used, section ("__OBJC, __cls_meth")))= {
  0, 1,
  {
    {
      (SEL)"doSomethingStatic",
      "v16@0:8",
      (void *)_C_SomeClass_doSomethingStatic
    }
  }	
};

static struct _objc_class _OBJC_METACLASS_SomeClass 
  __attribute__((used, section ("__OBJC, __meta_class")))= {
  (struct _objc_class *)"SomeClass",
  0,
  "SomeClass",
  0,
  2,
  sizeof(struct _objc_class),
  0, 
  (struct _objc_method_list *)&_OBJC_CLASS_METHODS_SomeClass,
  0,0,0,0
};

static struct _objc_class _OBJC_CLASS_SomeClass 
  __attribute__((used, section ("__OBJC, __class")))= {
  &_OBJC_METACLASS_SomeClass,
  0,
  "SomeClass",
  0, 1, 0, 0,
  (struct _objc_method_list *)&_OBJC_INSTANCE_METHODS_SomeClass, 
  0, 0, 0, 0
};

\end{verbatim}

There might be a lot of ``black magic`` going on. Note that this is not a real feature of clang compiler that is meant to be used in real life. It's more of a debugging tool. If you ever decide to programmatically create Objective-C classes, you definitely won't be creating structures like in the code above, but would call appropriate run-time methods for class allocations. As you can see here in the \verb=__attribute__= annotation, these structures will be stored in the \verb=__OBJC, *= sections of the resulting Mach-O binary image. This is quite system-specific, as I will mention later on when analyzing Apple's implementation of the run-time, there are many dyld\footnote{Apple's dynamic linker, open-source as a part of Darwin - http://www.opensource.apple.com/source/dyld/)}-specific functions that get called by the linker.

This saves time when loading binary images - the classes are already constructed and allocated, but is subject to change with each run-time version change.

Back to the generated code, though. As you can see, there are two generated functions that correspond to the methods declared by the class \verb=SomeClass=. Their names get slightly obfuscated - \verb=_X_ClassName_method_name_= - where X is either I for instance methods or C for class methods.

First two arguments are always the same - \verb=self= and \verb=_cmd=: \verb=self= points to the object the message is being sent and \verb=_cmd= is the selector (\verb=SEL=). Selector is a structure that consists of just the method's name, so theoretically, it's possible to simply retype \verb=char*= to \verb=SEL= as is being done later on in the method lists. Selectors have generally nothing to do with the actual method implementation.

The first two methods do nothing, hence are uninteresting, but the second one calls (sends a message) its own method \verb=secondMethod:withTwoArguments:=. There's a lot of retyping going on within the call, but generally a function called \verb=objc_msgSend= is called, which lies at the core of all message calls\footnote{There are actually specific functions for methods that return floating point numbers or structures, as these may require special ABI treatment on some architectures.}. The first argument of the function is just passing the \verb=self= variable, for the \verb=_cmd= argument, it uses result of \verb=sel_registerName= - the result is a \verb=SEL= and this method makes sure that for selectors with the same name, the same pointer is returned - i.e. two objects with a method with the same name get the same \verb=SEL= pointer. The third argument is quite self-explanatory.


\chapter{Apple's Implementation}
\section{Limitations}
// TODO
- Bridging, 16B objects minimum.

\end{document}
      